### 2019-06-13 2-Chapters/1-Chapter/chapter1.tex
Last proof reading of Ch6 and conclusion

The aim of this section is to assess the potential gain of learning algorithms in IoT scenarios, even when the number of intelligent devices in the network increases,
and the network usage is more and more fluctuatating.
% and the stochastic hypothesis is more and more violated.
To do that, we suppose an IoT network made of two types of devices: \emph{static devices} that use only one channel (fixed in time), and \emph{dynamic devices} that can choose the channel for each of their transmissions. Static devices form an interfering traffic, which could be generated by devices using other standards as well.
Note that instead of assuming that each static device uses a fixed channel, we could also assume a looser hypothesis: if each static device uses a fixed sub-set of the $K$ channels, and a purely uniform random access in its set of considered channels, then \emph{in average} the observed occupancy of the $K$ channels can be modeled as if it were occupied by (more) static devices using only one channel. So our first hypothesis is actually not constraining.
We first evaluate the probability of collision if dynamic devices randomly select channels (that is, a naive approach), and if a centralized controller would optimally distribute all of them in the channels at the beginning of the scenario.
This second approach is ideal, but not realistic the most common situation of decentralized co-located IoT networks, and it is just used here as a reference.
Then, these three reference scenarios allow to evaluate the performance of bandit algorithms, such as \UCB{} and TS, in a decentralized network, in terms of successful communication rate, as it reflects the network efficiency.
We show that these algorithms have near-optimal performance, even when the proportion of end-devices increases and the interfering traffic from other devices becomes less and less stochastic.

>>>>

The aim of this section is to assess the potential gain of learning algorithms in IoT scenarios, even when the number of intelligent devices in the network increases,
and the network usage is more and more fluctuating.
% and the stochastic hypothesis is more and more violated.
To do that, we suppose an IoT network made of two types of devices: \emph{static devices} that use only one channel (fixed in time), and \emph{dynamic devices} that can choose the channel for each of their transmissions. Static devices form an interfering traffic, which could be generated by devices using other standards as well.
Note that instead of assuming that each static device uses a fixed channel, we could also assume a looser hypothesis: if each static device uses a fixed sub-set of the $K$ channels, and a purely uniform random access in its set of considered channels, then \emph{in average} the observed occupancy of the $K$ channels can be modeled as if it were occupied by (more) static devices using only one channel. So our first hypothesis is actually not constraining.
We first evaluate the probability of collision if dynamic devices randomly select channels (that is, a naive approach), and if a centralized controller would optimally distribute all of them in the channels at the beginning of the scenario.
This second approach is ideal, but not realistic the most common situation of decentralized co-located IoT networks, and it is just used here as a reference.
Then, these three reference scenarios allow to evaluate the performance of bandit algorithms, such as \UCB{} and TS, in a decentralized network, in terms of successful communication rate, as it reflects the network efficiency.
We show that these algorithms have near-optimal performance, even when the proportion of end-devices increases and the interfering traffic from other devices becomes less and less stochastic.

### 2019-06-13 2-Chapters/2-Chapter/chapter2.tex
Done for now, I added a line about Hélène ❤
Last proof reading of Ch6 and conclusion

    The KL divergence between $B(x)$ and $B(y)$, of parameters $\eta$, $\eta'$ is given by
    $\kl(x,y) \eqdef \KL_{B}(\eta,\eta') = x \log(x/y) + (1-x) \log((1-x)/(1-y))$,
    and it is also called the \emph{relative binary entropy}.
    It satisfies $\kl(x,y) \geq d_{1/4}(x,y)$ (Pinsker's inadequality).

@@

\paragraph{Applications for Opportunistic Spectrum Access.}
% Detail the previous work from our team SCEE on bandits + OSA : Wassim, Navik

The focus of this work is on cognitive radio and IoT networks, where arms can represent wireless orthogonal channels, but more generally any resource characterizing the communication between a wireless device and a gateway (\eg, spreading factor for LoRa \cite{KerkoucheAlami18}, power allocation for NOMA etc). In cognitive radio using centralized supervision, for instance if the gateway can decide the allocation of devices to resources, MAB can also be used to let the gateway explore different allocations and learn by itself a good allocation, see for instance this article that consider 5G-like networks with small cells \cite{Maghsudi16}.

Previous works of our SCEE team showed that MAB can be used to model the problem of spectrum access for a secondary user accessing a licensed spectrum.
In this model, arms represent a finite set of orthogonal channels, \ie, different frequency bands in a licensed spectrum.
In the model with sensing, the samples $Y_{k,t}$ represents the feedback obtained by the CR-equipped device after sensing the channel $k$ at time $t$.
A reward of $r(t) = 1$ indicates that no Primary User was sensed, while a reward of $r(t)=0$ indicates that the channel $k$ is busy at time $t$ and no uplink message can be sent.
%
This model was first introduced by Wassim Jouini during his PhD thesis,
in \cite{Jouini09} and later studied in both \cite{Jouini10,Jouini12}.
Proof-of-concepts using real-world radio hardware was first propsed in \cite{MoyWSR2014,RobertSDR2014}.

\TODOL{Je sais pas à quel point je veux parler de la thèse de Navik? Peut-être que j'en aurai dit assez au Chapitre 1, et que là ça suffira..}
In a second PhD thesis supervised by Christophe Moy, Navikkumar Modi studied the impact
of using MAB algorithms to optimize channel selection
on the battery life of a wireless device.
On the one hand, running a MAB algorithm such as UCB-like algorithms was proven to be useful and can bring significant improvement in terms of successful transmission rates, directly increasing the battery life of the device.
On the other hand, classical MAB algorithms tend to switch arms a lot of times, especially in the beginning of the learning process, and this induces a lot of dynamical hardware reconfiguration for the wireless device, as selecting a different channel requires a change in the radio hardware used by the device.
Each hardware reconfiguration costs energy for the device, and quickly switching algorithms will lead to a reduction of the battery life.
The tradeoff between the two aspects is studied empirically in
\cite{modiDemo2016}.
%
Another interesting works is \cite{Modi17QoS}.

\TODOL{FIXME Finir ce blabla rapidement ? Ou alors ce sera déjà fait en intro/chapitre1 ??}

@@

% - Worst case regret (max $R_T^{r}$ for $r$ index of Monte-Carlo simulation?)
The \textbf{switching cost} $\mathrm{SC}^{\cA}(T)$ counts how many times the player's decision has changed from one time step to the next one, \ie, $\mathrm{SC}^{\cA}(T) \eqdef \sum_{t=1}^{T-1} \indic(A(t) \neq A(t+1))$.
It has recently gained interest in the literature, for instance the authors of \cite{Koren17} proposed an algorithm that adaptatively tries to balance the trade-off between minimizing the regret and minimizing the switching cost.
%
Indeed, in single-player models it is easy to show that achieving logarithmic regret directly implies a logarithmic upper-bound on the switching cost, and conversely the lower-bound from \cite{LaiRobbins85} also gives an asymptotic logarithmic lower-bound.
But even if $\mathrm{SC}^{\cA}(T) = \Theta(\log T)$ for an efficient algorithm, it can be interesting to numerically evaluate this quantity, as a large value might indicate an algorithm that is alternating too much between the optimal arm and other arms.

\TODOL{Enlever ça, juste une phrase qui dit "on le fait plus tard".}
We find more interesting to study $\mathrm{SC}^{\cA_1,\dots,\cA_M}(T)$, the sum of the switching costs of the $M$ players in a multi-players bandit game (if player $j$ uses algorithm $\cA_j$ for $j=1,\dots,M$). Indeed in this case it is possible that the players attain a logarithmic centralized regret, by following an efficient centralized or decentralized algorithm, while still suffering from a linear switching cost.
%
It is very satisfying to obtain a logarithmic switching cost for our algorithm \MCTopM-\klUCB{} in Chapter~\ref{chapter:5}, even if minimizing the switching cost was not one of our objective, and the bound we obtain on $\mathrm{SC}^{\cA_1,\dots,\cA_M}(T)$ is just a consequence of our proof of the control we give on the regret $R^{\cA}(T)$.
% - Switching cost? \cite{modiDemo2016} \cite{Koren17}
On a more experimental note, it was shown in \cite{modiDemo2016} that the previous state-of-the-art policy for multi-players bandits, \rhoRand, could be tuned to run in batches\footnote{~The batch bandit setting implies that all players use the same decisions for instance for $50$ consecutive times, and update their decisions only once every $50$ time steps. For more details, see \cite{modiDemo2016} for experiments on the multi-players case, or \cite{perchet2016,gao2019batched,kolnogorov2019multi} for theoretical developments on the single-player case.}, in order to reduce by a certain multiplicative factor its switching cost while only adding an additive factor on its regret. While such ideas can interest a practitioner, they does not change the asymptotic behavior of $\mathrm{SC}^{\cA_1,\dots,\cA_M}(T)$, which is $\Theta(\log(T))$ for any efficient policy.

% - Fairness ?
Finally, another interesting measure of performance is the \textbf{fairness}.
It was not studied much for single-player problems, and usually it means that each arm must be explored at least a given fraction of the total horizon $T$ (\ie, ``arm-wise fairness'').
It was studied in a very recent article \cite{Patil2019stochastic},
where a different notion of regret is introduced to account for the fairness constraint. The authors show that different algorithms, based on $\UCB_1$ or Thompson sampling, can achieve logarithmic regret while respecting the fairness constraints.
%
\TODOL{Enlever ça, juste une phrase qui dit "on le fait plus tard".}
In this thesis, we only consider fairness in the multi-players bandit models,
in Chapter~\ref{chapter:5}, where fairness refers to a different notion (\ie, ``cooperative fairness'').
When $M<K$ players learn in a decentralized way, they will essentially converge to an orthogonal affectation of players to arms, so that each player uses a unique arm among the $M$ best arms (most of the times).
The fairness constraint says that, in average, no user is arbitrarily going to converge more frequently on a better arm than another user, and it is for instance the case for permutation invariant decentralized algorithms.
We refer to Section~\ref{sub:5:betterLowerBound} for more details.

@@

\textbf{The $\UCB_1$ index policy: using Hoeffding's inequality to build confidence intervals.}\\
%
Let us consider another approach for adaptive exploration, known as ``optimism under uncertainty'': assume each arm is as good as it can possibly be given the observations so far, and choose the best arm based on these optimistic estimate.
%
This intuition leads to the $\UCB_1$ algorithm, initially introduced in \cite{Auer02}.
For a parameter $\alpha$, the \UCB{} indexes are computed as follows, as a sum of
the \emph{average reward} $\widehat{\mu_k}(t) \eqdef \frac{X_k(t)}{N_k(t)}$
and a \emph{confidence radius} $\xi_k(t) \eqdef \sqrt{\alpha \frac{\log\left(t\right)}{N_k(t)}}$.
%
\begin{align}\label{eq:UCB_index}
    U_k^{\UCB_1}(t) = \UCB_k(t) \eqdef \underbrace{\frac{X_k(t)}{N_k(t)}}_{\text{Exploitation }\; \widehat{\mu_k}(t)} + \underbrace{\sqrt{\alpha \frac{\log\left(t\right)}{N_k(t)}}}_{\text{Exploration }\; \xi_k(t)}.
\end{align}

This selection rule $A(t) \in \cU(\arg\max\limits_{1\leq k \leq K} U_k^{\cA}(t))$ makes sense for the following reasons:
an arm $k$ is chosen at step $t$ because it has a large $\UCB_k(t)$, which can happen for two reasons.
$i)$ Because the average reward $\widehat{\mu_k}(t)$ is large, in which case this arm is likely to have a high mean reward $\mu_k$,
$ii)$ and/or because the confidence radius $\xi_k(t)$ is large, in which case this arm has not been explored much.
Either reason makes this arm worth choosing.
One can also observe that $\xi_k(t)\to\infty$ if $t\to\infty$ while $N_k(t)\ll\log(t)$, which can be used to prove that the $\UCB_1$ algorithm tries all arms at least $\Omega(\log(T))$ times (in average).
%
Furthermore, these two terms $\widehat{\mu_k}(t)$ and $\xi_k(t)$ in the \UCB{} defined in \eqref{eq:UCB_index} respectively represent \emph{exploitation} and \emph{exploration}, and summing them up is a natural way of dealing with the exploitation/exploration trade off.
The constant parameter $\alpha$ in $\xi_k(t)$ also controls this trade-off, and the theoretical analysis suggests to restrict to $\alpha\geq1/2$, and to choose $\alpha=1/2$ for optimal uniform performance (\ie, on all problems, \cite{Auer02}).

@@

Instead of using a confidence radius $\xi_k(t)$ based on Hoeffding's inequality,
the \klUCB{} algorithm rather use the Chernoff concentration inequality for the considered exponential family
% such as Chernoff-Hoeffding for Gaussian (or sub-Gaussian) rewards.
% It yields improved upper confidence bounds (\ie, with smaller $\xi_k(t)$) that are based on Kullback-Leibler divergence.
The indexes are computed as follows, for an exploration function $f(t)$
that is typically chosen as $f(t) \eqdef \log(t)+c\log(\log(t))$ for the theoretical analyses (with $c \geq 3$),
and $f(t) \eqdef \log(t)$ in practice.
The constant $c>0$ is usually chosen as $c = 1$.
\begin{align}\label{eq:klUCB_index}
    U_k^{\klUCB}(t) \eqdef \sup\limits_{q \in [0, 1]} \left\{ q : \kl\left(\frac{X_k(t)}{N_k(t)}, q\right) \leq \frac{f(t}{N_k(t)} \right\}.
\end{align}

@@

\paragraph{A note about the implementation of \klUCB{} algorithms.}
%
Implementing this kind of policy requires to implement two things.
The first piece is usually easy: one has to implement a way to compute $\kl(x,y)$ the Kullback-Leibler divergence of the considered one-dimensional family,
for instance $\kl$ for Bernoulli distributions has a closed form and is easy to compute.
The second piece is less obvious: one has to solve the optimisation problem defining the $\sup$ in \eqref{eq:klUCB_index}.
Any iterative algorithm for numerical optimization of one-dimensional functions can be used (\eg, see \cite{BoydVanderberghe04} for a survey), but as the $\kl(x,\bullet)$ function is usually convex (for any $x$), two practical and mathematically correct solutions are Newton's method or a bisection search.
The search space for $q$ can be restricted to $[ X_k(t) / N_k(t), 1 ]$ instead of $[0,1]$,
and for simplicity and efficiency, in this case a naive bisection search works well.
% and Newton's method guarantees that a small number of steps yield a given precision level $\varepsilon$.
For instance, $\varepsilon=10^{-8}$ is enough for numerical simulations, and typically the bisection search terminates in between $5$ to $20$ steps.

@@

For curiosity, we mention a third Bayesian algorithm, AdBandits introduced in \cite{Truzzi13}, mainly because it has not gained any popularity despite its good empirical performance.
%  using $\alpha=1$ (\texttt{AdBandits}).
AdBandits follows the same assumptions as Thompson sampling and Bayes-UCB, and for instance for Bernoulli problems it uses a Beta posterior on each arm.
It uses a random mixture between the two algorithms, and with a certain probability it uses the Thompson sampling decision rule, otherwise it uses the Bayes-UCB decision rule.
It was proposed in this article \cite{Truzzi13} and illustrated on empirical simulations, but no theoretical analysis was given.
We found that in practice it can achieve very good performance, as illustrated in Section~\ref{sec:3:reviewSPAlgorithms}.
%  and it did not gain popularity since then.
% In practice, AdBandits is usually (slighly) more efficient than both Thompson sampling and Bayes-UCB in terms of regret, it is comparable in terms of computation times, but costs more memory.

@@

% ----------------------------------------------------------------------------
\textbf{Policies for adversarial bandits.}
%
Another important family of bandits algorithms are the algorithms proposed for the adversarial setting \cite{Auer02NonStochastic}.
The first of them is the Exp3 policy, for ``\emph{Exp}onential weights for \emph{Exp}loitation and \emph{Exp}loiration'',
which is an example of the well-known \emph{multiplicative weights update algorithm}\footnote{~For more details see
% on Wikipedia \href{https://en.wikipedia.org/wiki/Multiplicative_weight_update_method}{\texttt{en.wikipedia.org/wiki/Multiplicative\_weight\_update\_method}},
this blog post by Jeremy Kun, \href{https://jeremykun.com/2017/02/27/the-reasonable-effectiveness-of-the-multiplicative-weights-update-algorithm/}{\texttt{JeremyKun.com/2017/02/27/the-reasonable-}}\\
\href{https://jeremykun.com/2017/02/27/the-reasonable-effectiveness-of-the-multiplicative-weights-update-algorithm/}{\texttt{effectiveness-of-the-multiplicative-weights-update-algorithm}},
which comes with an interactive online demonstration to help understanding the behavior of the algorithm, hosted at \href{https://j2kun.github.io/mwua/index.html}{\texttt{j2kun.GitHub.io/mwua/}} (accessed the 24th of May 2019). More details can be found in the survey \cite{Arora2012multiplicative}.},
a family of algorithms developed since the 1950s, and applied to a wide range of different domains.
% https://smpybandits.github.io/docs/Policies.Exp3.html
%
The Exp3 policy maintains weights $w_k(t)$ on the $K$ arms, and at time $t$ it samples an arm from the distribution which is a mixture of $\bm{w}(t)=[w_1(t),\dots,w_K(t)]$ and the uniform distribution, \ie,
$\Pr(A(t)=k) \eqdef (1-\gamma) w_k(t) + \gamma \frac{1}{K}$.
The weights are initially uniform, \ie, $w_k(0)=1/K$, and then at each time after observing a reward $Y_{k,t}$ from arm $k$, Exp3 updates the weight of the chosen arm multiplicatively, using $w_k(t+1) = w_k(t) \times \exp(\widehat{r_k}(t) / (\gamma N_k(t)))$. The weights are then renormalized, to have a sum $\sum_k w_k(t+1) = 1$.
This notation $\widehat{r_k}(t)$ means an unbiased estimator of the reward $r(t)=Y_{k,t}$, that is, $\widehat{r_k}(t) = r(t) / \Pr(A(t)=k)$.
%
Like for the $\varepsilon$-greedy policy, the parameter $\gamma$ can be constant, or can be a non-increasing sequence $(\gamma_t)_{t\in\N^*}$.
%
It was proven in \cite{Auer02NonStochastic} that Exp3 with constant parameter $\gamma$ achieves $R_T = \bigO{\sqrt{K T \log(T)}}$ problem-independent regret,
while using a decreasing sequence, such as $\gamma_t = \sqrt{\log(K) / K} \log(t) / t$ gives an order-optimal problem-independent regret upper-bound of $\bigO{\sqrt{K T}}$.
A short proof of this results and more details on Exp3 are given in \cite{Bubeck12}.
%
Many other policies have been proposed for adversarial bandits, like Follow-the-Perturbed-Leader or Follow-the-Regularized-Leader for linear adversarial bandits,
% but we do not study them in this thesis,
and we refer to Parts~III and VI of \cite{LattimoreBanditAlgorithmsBook} for more details.

@@

If the two indexes $U^{\KLUCB+}_k(t)$ and $U^{\mathrm{MOSS}+}_k(t)$ are defined by

% \vspace*{-20pt}   % FIXME??
\begin{align*}
    U^{\mathrm{MOSS}+}_k(t) & \eqdef \frac{X_k(t)}{N_k(t)} + \max\left(0, \sqrt{ \frac{\log\left(\frac{T}{K N_k(t)}\right)}{N_k(t)}} \right). \\
% \end{align*}
% \begin{align*}
    U^{\KLUCB+}_k(t) & \eqdef \sup\limits_{q \in [a, b]} \left\{ q : \kl\left(\frac{X_k(t)}{N_k(t)}, q\right) \leq \frac{\log(\frac{T}{K N_k(t)})}{N_k(t)} \right\} \\
% \end{align*}
% \begin{align*}
    % U^{\klUCB\text{-}\mathrm{switch}}_k(t) & \eqdef \begin{cases}
    %     U^{\KLUCB+}_k(t) & \text{if } N_k(t) \leq f(T, K), \\
    %     U^{\mathrm{MOSS}+}_k(t) & \text{if } N_k(t) > f(T, K).
    % \end{cases}.
\end{align*}

>>>>

    The KL divergence between $B(x)$ and $B(y)$, of parameters $\eta$, $\eta'$ is given by
    $\kl(x,y) \eqdef \KL_{B}(\eta,\eta') = x \log(x/y) + (1-x) \log((1-x)/(1-y))$,
    and it is also called the \emph{relative binary entropy}.
    It satisfies $\kl(x,y) \geq d_{1/4}(x,y)$ (Pinsker's inequality).

@@

\textbf{Applications for Opportunistic Spectrum Access.}
% Detail the previous work from our team SCEE on bandits + OSA : Wassim, Navik
%
As highlighted in Chapter~\ref{chapter:1},
the focus of this work is on cognitive radio and IoT networks, where arms can represent wireless orthogonal channels, but more generally any resource characterizing the communication between a wireless device and a gateway (\eg, spreading factor for LoRa \cite{KerkoucheAlami18}, power allocation for NOMA etc).
% In cognitive radio using centralized supervision, for instance if the gateway can decide the allocation of devices to resources, MAB can also be used to let the gateway explore different allocations and learn by itself a good allocation, see for instance this article that consider 5G-like networks with small cells \cite{Maghsudi16}.
%
% Previous works of our SCEE team showed that MAB can be used to model the problem of spectrum access for a secondary user accessing a licensed spectrum.
% In this model, arms represent a finite set of orthogonal channels, \ie, different frequency bands in a licensed spectrum.
In the model of Opportunistic Spectrum Access (OSA) with sensing, the samples $Y_{k,t}$ represents the feedback obtained by the CR-equipped device after sensing the channel $k$ at time $t$.
A reward of $r(t) = 1$ indicates that no Primary User was sensed, while a reward of $r(t)=0$ indicates that the channel $k$ is busy at time $t$ and no uplink message can be sent.
%
This model was first introduced by Wassim Jouini during his PhD thesis,
in \cite{Jouini09} and later studied in both \cite{Jouini10,Jouini12}.
Proof-of-concepts using real-world radio hardware was first proposed in \cite{MoyWSR2014,RobertSDR2014}.
%
We study a similar model of using MAB for Cognitive Radio but without the OSA strucutre of Primary and Secondary users, \ie, without sensing information but only a collision indicator, in Chapters~\ref{chapter:4} and \ref{chapter:5}.

@@

% - Worst case regret (max $R_T^{r}$ for $r$ index of Monte-Carlo simulation?)
The \textbf{switching cost} $\mathrm{SC}^{\cA}(T)$ counts how many times the player's decision has changed from one time step to the next one, \ie, $\mathrm{SC}^{\cA}(T) \eqdef \sum_{t=1}^{T-1} \indic(A(t) \neq A(t+1))$.
It has recently gained interest in the literature, for instance the authors of \cite{Koren17} proposed an algorithm that adaptively tries to balance the trade-off between minimizing the regret and minimizing the switching cost.
%
Indeed, in single-player models it is easy to show that achieving logarithmic regret directly implies a logarithmic upper-bound on the switching cost, and conversely the lower-bound from \cite{LaiRobbins85} also gives an asymptotic logarithmic lower-bound.
But even if $\mathrm{SC}^{\cA}(T) = \Theta(\log T)$ for an efficient algorithm, it can be interesting to numerically evaluate this quantity, as a large value might indicate an algorithm that is alternating too much between the optimal arm and other arms.
%
% \TODOL{Enlever ça, juste une phrase qui dit "on le fait plus tard".}
We find more interesting to study $\mathrm{SC}^{\cA_1,\dots,\cA_M}(T)$, the sum of the switching costs of the $M$ players in a multi-players bandit game (if player $j$ uses algorithm $\cA_j$ for $j=1,\dots,M$), as studied in Chapter~\ref{chapter:5}.
% Indeed in this case it is possible that the players attain a logarithmic centralized regret, by following an efficient centralized or decentralized algorithm, while still suffering from a linear switching cost.
% %
% It is very satisfying to obtain a logarithmic switching cost for our algorithm \MCTopM-\klUCB{} in Chapter~\ref{chapter:5}, even if minimizing the switching cost was not one of our objective, and the bound we obtain on $\mathrm{SC}^{\cA_1,\dots,\cA_M}(T)$ is just a consequence of our proof of the control we give on the regret $R^{\cA}(T)$.
% % - Switching cost? \cite{modiDemo2016} \cite{Koren17}
% On a more experimental note, it was shown in \cite{modiDemo2016} that the previous state-of-the-art policy for multi-players bandits, \rhoRand, could be tuned to run in batches\footnote{~The batch bandit setting implies that all players use the same decisions for instance for $50$ consecutive times, and update their decisions only once every $50$ time steps. For more details, see \cite{modiDemo2016} for experiments on the multi-players case, or \cite{perchet2016,gao2019batched,kolnogorov2019multi} for theoretical developments on the single-player case.}, in order to reduce by a certain multiplicative factor its switching cost while only adding an additive factor on its regret. While such ideas can interest a practitioner, they does not change the asymptotic behavior of $\mathrm{SC}^{\cA_1,\dots,\cA_M}(T)$, which is $\Theta(\log(T))$ for any efficient policy.

% - Fairness ?
Finally, another interesting measure of performance is the \textbf{fairness}.
It was not studied much for single-player problems, and usually it means that each arm must be explored at least a given fraction of the total horizon $T$ (\ie, ``arm-wise fairness'').
It was studied in a very recent article \cite{Patil2019stochastic},
where a different notion of regret is introduced to account for the fairness constraint. The authors show that different algorithms, based on $\UCB_1$ or Thompson sampling, can achieve logarithmic regret while respecting the fairness constraints.
%
% \TODOL{Enlever ça, juste une phrase qui dit "on le fait plus tard".}
In this thesis, we only consider fairness in the multi-players bandit models,
in Chapter~\ref{chapter:5}, where fairness refers to a different notion (\ie, ``cooperative fairness'').
% When $M<K$ players learn in a decentralized way, they will essentially converge to an orthogonal affectation of players to arms, so that each player uses a unique arm among the $M$ best arms (most of the times).
% The fairness constraint says that, in average, no user is arbitrarily going to converge more frequently on a better arm than another user, and it is for instance the case for permutation invariant decentralized algorithms.
We refer to Section~\ref{sub:5:betterLowerBound} for more details.

@@

\textbf{The $\UCB_1$ index policy: using Hoeffding's inequality to build confidence intervals.}\\
%
Let us consider another approach for adaptive exploration, known as ``optimism under uncertainty'': assume each arm is as good as it can possibly be given the observations so far, and choose the best arm based on these optimistic estimate.
%
This intuition leads to the $\UCB_1$ algorithm, initially introduced in \cite{Auer02}.
For a parameter $\alpha$, the \UCB{} indexes are computed as follows, as a sum of
the \emph{average reward} $\widehat{\mu_k}(t) \eqdef \frac{X_k(t)}{N_k(t)}$
and a \emph{confidence radius} $\xi_k(t) \eqdef \sqrt{\alpha \frac{\log\left(t\right)}{N_k(t)}}$.
%
\begin{align}\label{eq:2:UCB_index}
    U_k^{\UCB_1}(t) = \UCB_k(t) \eqdef \underbrace{\frac{X_k(t)}{N_k(t)}}_{\text{Exploitation }\; \widehat{\mu_k}(t)} + \underbrace{\sqrt{\alpha \frac{\log\left(t\right)}{N_k(t)}}}_{\text{Exploration }\; \xi_k(t)}.
\end{align}

This selection rule $A(t) \in \cU(\arg\max\limits_{1\leq k \leq K} U_k^{\cA}(t))$ makes sense for the following reasons:
an arm $k$ is chosen at step $t$ because it has a large $\UCB_k(t)$, which can happen for two reasons.
$i)$ Because the average reward $\widehat{\mu_k}(t)$ is large, in which case this arm is likely to have a high mean reward $\mu_k$,
$ii)$ and/or because the confidence radius $\xi_k(t)$ is large, in which case this arm has not been explored much.
Either reason makes this arm worth choosing.
One can also observe that $\xi_k(t)\to\infty$ if $t\to\infty$ while $N_k(t)\ll\log(t)$, which can be used to prove that the $\UCB_1$ algorithm tries all arms at least $\Omega(\log(T))$ times (in average).
%
Furthermore, these two terms $\widehat{\mu_k}(t)$ and $\xi_k(t)$ in the \UCB{} defined in \eqref{eq:2:UCB_index} respectively represent \emph{exploitation} and \emph{exploration}, and summing them up is a natural way of dealing with the exploitation/exploration trade off.
The constant parameter $\alpha$ in $\xi_k(t)$ also controls this trade-off, and the theoretical analysis suggests to restrict to $\alpha\geq1/2$, and to choose $\alpha=1/2$ for optimal uniform performance (\ie, on all problems, \cite{Auer02}).

@@

Instead of using a confidence radius $\xi_k(t)$ based on Hoeffding's inequality,
the \klUCB{} algorithm rather use the Chernoff concentration inequality for the considered exponential family
% such as Chernoff-Hoeffding for Gaussian (or sub-Gaussian) rewards.
% It yields improved upper confidence bounds (\ie, with smaller $\xi_k(t)$) that are based on Kullback-Leibler divergence.
The indexes are computed as follows, for an exploration function $f(t)$
that is typically chosen as $f(t) \eqdef \log(t)+c\log(\log(t))$ for the theoretical analyses (with $c \geq 3$),
and $f(t) \eqdef \log(t)$ in practice.
The constant $c>0$ is usually chosen as $c = 1$.
\begin{align}\label{eq:2:klUCB_index}
    U_k^{\klUCB}(t) \eqdef \sup\limits_{q \in [0, 1]} \left\{ q : \kl\left(\frac{X_k(t)}{N_k(t)}, q\right) \leq \frac{f(t}{N_k(t)} \right\}.
\end{align}

@@

\paragraph{A note about the implementation of \klUCB{} algorithms.}
%
Implementing this kind of policy requires to implement two things.
The first piece is usually easy: one has to implement a way to compute $\kl(x,y)$ the Kullback-Leibler divergence of the considered one-dimensional family,
for instance $\kl$ for Bernoulli distributions has a closed form and is easy to compute.
The second piece is less obvious: one has to solve the optimisation problem defining the $\sup$ in \eqref{eq:2:klUCB_index}.
Any iterative algorithm for numerical optimization of one-dimensional functions can be used (\eg, see \cite{BoydVanderberghe04} for a survey), but as the $\kl(x,\bullet)$ function is usually convex (for any $x$), two practical and mathematically correct solutions are Newton's method or a bisection search.
The search space for $q$ can be restricted to $[ X_k(t) / N_k(t), 1 ]$ instead of $[0,1]$,
and for simplicity and efficiency, in this case a naive bisection search works well.
% and Newton's method guarantees that a small number of steps yield a given precision level $\varepsilon$.
For instance, $\varepsilon=10^{-8}$ is enough for numerical simulations, and typically the bisection search terminates in between $5$ to $20$ steps.

@@

For curiosity, we mention a third Bayesian algorithm, AdBandits introduced in \cite{Truzzi13}, mainly because it has not gained any popularity despite its good empirical performance.
%  using $\alpha=1$ (\texttt{AdBandits}).
AdBandits follows the same assumptions as Thompson sampling and Bayes-UCB, and for instance for Bernoulli problems it uses a Beta posterior on each arm.
It uses a random mixture between the two algorithms, and with a certain probability it uses the Thompson sampling decision rule, otherwise it uses the Bayes-UCB decision rule.
It was proposed in this article \cite{Truzzi13} and illustrated on empirical simulations, but no theoretical analysis was given.
We found that in practice it can achieve very good performance, as illustrated in Section~\ref{sec:3:reviewSPAlgorithms}.
%  and it did not gain popularity since then.
% In practice, AdBandits is usually (slightly) more efficient than both Thompson sampling and Bayes-UCB in terms of regret, it is comparable in terms of computation times, but costs more memory.

@@

% ----------------------------------------------------------------------------
\textbf{Policies for adversarial bandits.}
%
Another important family of bandits algorithms are the algorithms proposed for the adversarial setting \cite{Auer02NonStochastic}.
The first of them is the Exp3 policy, for ``\emph{Exp}onential weights for \emph{Exp}loitation and \emph{Exp}loration'',
which is an example of the well-known \emph{multiplicative weights update algorithm}\footnote{~For more details see
% on Wikipedia \href{https://en.wikipedia.org/wiki/Multiplicative_weight_update_method}{\texttt{en.wikipedia.org/wiki/Multiplicative\_weight\_update\_method}},
this blog post by Jeremy Kun, \href{https://jeremykun.com/2017/02/27/the-reasonable-effectiveness-of-the-multiplicative-weights-update-algorithm/}{\texttt{JeremyKun.com/2017/02/27/the-reasonable-}}\\
\href{https://jeremykun.com/2017/02/27/the-reasonable-effectiveness-of-the-multiplicative-weights-update-algorithm/}{\texttt{effectiveness-of-the-multiplicative-weights-update-algorithm}},
which comes with an interactive online demonstration to help understanding the behavior of the algorithm, hosted at \href{https://j2kun.github.io/mwua/index.html}{\texttt{j2kun.GitHub.io/mwua/}} (accessed the 24th of May 2019). More details can be found in the survey \cite{Arora2012multiplicative}.},
a family of algorithms developed since the 1950s, and applied to a wide range of different domains.
% https://smpybandits.github.io/docs/Policies.Exp3.html
%
The Exp3 policy maintains weights $w_k(t)$ on the $K$ arms, and at time $t$ it samples an arm from the distribution which is a mixture of $\bm{w}(t)=[w_1(t),\dots,w_K(t)]$ and the uniform distribution, \ie,
$\Pr(A(t)=k) \eqdef (1-\gamma) w_k(t) + \gamma \frac{1}{K}$.
The weights are initially uniform, \ie, $w_k(0)=1/K$, and then at each time after observing a reward $Y_{k,t}$ from arm $k$, Exp3 updates the weight of the chosen arm multiplicatively, using $w_k(t+1) = w_k(t) \times \exp(\widehat{r_k}(t) / (\gamma N_k(t)))$. The weights are then renormalized, to have a sum $\sum_k w_k(t+1) = 1$.
This notation $\widehat{r_k}(t)$ means an unbiased estimator of the reward $r(t)=Y_{k,t}$, that is, $\widehat{r_k}(t) = r(t) / \Pr(A(t)=k)$.
%
Like for the $\varepsilon$-greedy policy, the parameter $\gamma$ can be constant, or can be a non-increasing sequence $(\gamma_t)_{t\in\N^*}$.
%
It was proven in \cite{Auer02NonStochastic} that Exp3 with constant parameter $\gamma$ achieves $R_T = \bigO{\sqrt{K T \log(T)}}$ problem-independent regret,
while using a decreasing sequence, such as $\gamma_t = \sqrt{\log(K) / K} \log(t) / t$ gives an order-optimal problem-independent regret upper-bound of $\bigO{\sqrt{K T}}$.
A short proof of this results and more details on Exp3 are given in \cite{Bubeck12}.
%
Many other policies have been proposed for adversarial bandits, like Follow-the-Perturbed-Leader or Follow-the-Regularized-Leader for linear adversarial bandits,
% but we do not study them in this thesis,
and we refer to Parts~III and VI of \cite{LattimoreBanditAlgorithmsBook} for more details.

@@

If the two indexes $U^{\KLUCB+}_k(t)$ and $U^{\mathrm{MOSS}+}_k(t)$ are defined by
%
% \vspace*{-20pt}   % FIXME??
\begin{align*}
    U^{\mathrm{MOSS}+}_k(t) & \eqdef \frac{X_k(t)}{N_k(t)} + \max\left(0, \sqrt{ \frac{\log\left(\frac{T}{K N_k(t)}\right)}{N_k(t)}} \right). \\
% \end{align*}
% \begin{align*}
    U^{\KLUCB+}_k(t) & \eqdef \sup\limits_{q \in [a, b]} \left\{ q : \kl\left(\frac{X_k(t)}{N_k(t)}, q\right) \leq \frac{\log(\frac{T}{K N_k(t)})}{N_k(t)} \right\}
% \end{align*}
% \begin{align*}
    % U^{\klUCB\text{-}\mathrm{switch}}_k(t) & \eqdef \begin{cases}
    %     U^{\KLUCB+}_k(t) & \text{if } N_k(t) \leq f(T, K), \\
    %     U^{\mathrm{MOSS}+}_k(t) & \text{if } N_k(t) > f(T, K).
    % \end{cases}.
\end{align*}
